A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}

### 2020-07-28 article_format/title.tex
Changed the formating, made some minor edits

% see https://stackoverflow.com/a/47122900
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

\newpage

\pagenumbering{arabic} 
\setcounter{page}{1}

\doublespacing

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Several years ago, \citet{Curran2011} reflected positively on the
growing use of panel studies in empirical social research. Some of the
strengths of panel data are well-known, e.g., the ability to establish
temporal precedence, increased statistical power and the reduction of
potential alternative models. However, perhaps the greatest strength of
panel data is that they allow for a more rigorous testing of substantive
theories. Panel data, i.e., repeated measures of the same observed units
(people, schools, firms, countries, etc.), allow researchers to
decompose the error term into a part that stays constant within units
and the part that changes over time. The part that does not change over
time can be seen as the combined effect of all time-invariant influences
(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
effects (FE) regression involves controlling for these time-invariant
influences via a number of various methods. It thus accounts for a
likely and common source of bias.

Structural equation modeling (SEM) is a popular regression framework.
One of its main strengths is its flexibility. Not only can complex
causal structures with multiple dependent variables be tested
simultaneously, but in longitudinal (and, more generally, hierarchical)
studies both time-varying and invariant predictors can be included, and
effects can easily be allowed to vary over time. Thus researchers can
allow for and study effects that increase or fade over time, or that
appear only in specific periods. Beyond that, with the use of latent
variables, SEM provides a way to deal with measurement error and get
closer to the true underlying constructs of interest.

There are a number of articles describing basic concept of panel model
regression, and FE regression in SEM
\citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
intended as a \textit{practical guide} for researchers looking for
in-depth help with specifying FE models in SEM. It focuses on the
\texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
While \texttt{Mplus} \citep{Mplus} is arguably the most robust SEM
software currently available (in terms of features like alignment,
latent variable interactions, for example), the \texttt{lavaan} package
has many benefits. First, like \texttt{R} it is open source and
completely free. For researchers dipping their toes into SEM, there is
no financial barrier to try, and no risk if they decide it is not for
them. Second, the implementation of \texttt{lavaan} in the larger
\texttt{R} environment is an enormous advantage. Instead of poring over
reams of plain text, copying out coefficients by hand, every part of the
\texttt{lavaan} output is available as an object. This means that all
aspects of the model, from fit indices, to coefficients and standard
errors, to the model matrices, can be accessed and easily integrated
into tables and plots. Furthermore, \texttt{R} can be used for a great
deal of applications. It can be used to manage and manipulate as well as
simulate data, perform symbolic algebra, run more traditional analyses
(e.g., multiple regression, logistic regression, principal component
analysis), etc. Once one is comfortable using \texttt{R}, there is no
longer any need to switch between different software for data
preparation and analysis.

The following article outlines the basic idea of panel regression, the
particularities of panel regression SEM, and shows its implementation in
\texttt{lavaan}. The focus on FE panel regression stems from the fact
that the assumptions needed to justify other common models, e.g., random
effects, are often implausible. As we will see, the practical
differences between the two are, in any event, very small. Using
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
data}, it demonstrates and annotates the code for the most basic FE
model and provides an overview of the summary output. One of the main
strengths of panel SEM compared to the more traditional methods of panel
analysis is its flexibility. Therefore, a number of potential extensions
to the basic model, including relaxing various assumptions, dealing with
measurement error in both the independent and dependent variables, as
well as the inclusion of time-invariant predictors in the form of a
hybrid fixed-/ random effects model, are shown in detail in the form of
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials}.

\hypertarget{panel}{%
\section{Panel models}\label{panel}}

To begin, let us start by reviewing a general panel model
\citep{Bollen2010}, also referred to as the `unobserved effects model'
\citep{Wooldridge2012, R-plm_a} (we will return to this model in the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials} when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align} where \(y_{it}\) is the dependent variable for unit
\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
is a \(1 \times K\) vector of time-varying covariates (which could
include a constant) linked to the dependent variable by the
\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
a \(1 \times M\) vector of time-invariant covariates linked to the
dependent variable by the \(M \times 1\) vector of coefficients in
\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
unobserved time-constant variables affecting the dependent variable and
\(\varepsilon_{it}\) is the idiosyncratic error.

We can make stating some of the model assumptions easier by rewriting it
in matrix notation \begin{align}
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\iota}_{T}\alpha_{i} + \bm{\varepsilon}_{i}
\end{align} where \(\bm{y}_{i}\) and \(\bm{\varepsilon}_{i}\) are
\(T \times 1\) vectors, \(\bm{X}_{i}\) and \(\bm{Z}_{i}\) are
\(T \times K\) and \(T \times M\) matrices, respectively,
\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones and \(\alpha_{i}\)
is a scalar. \(\bm{\beta}\) and \(\bm{\gamma}\) are unchanged from
Equation \eqref{eq:gpm}.

Consistency of the following models requires the assumption of strict
exogeneity, although what constitutes strict exogeneity differs between
the random and fixed effects setups. Each assumption will be discussed
shortly. Apart from that, we typically make the following assumptions
about this model \citep[see, e.g.,][]{Wooldridge2002, Schmidheiny2019}:

\begin{itemize}
\tightlist
\item
  Linearity: the model is linear in its parameters.
\item
  Independence: the observations are independent across individuals
  (assured by random sampling in the cross-section), but not necessarily
  across time.
\item
  The usual rank condition: we have more observations than independent
  variables and there is no perfect collinearity between any of the
  independent variables.
\end{itemize}

\hypertarget{re}{%
\subsection{Random effects}\label{re}}

Somewhat unintuitively, the modern approach to FE regression treats the
unobserved effect \(\alpha_{i}\) as a \textit{random} variable, rather
than a fixed parameter to be estimated for each of the cross sectional
units \citep{Wooldridge2002}. And in fact, this view carries over to the
way in which FE regression is applied in SEM. So, in order to help
facilitate the discussion on the application of FE-SEM later on, it
makes sense to spend some time discussing the basic idea of random
effects regression (again, as we will see, the practical difference
between the two models is very small).

For the random effects (RE) model, we define a composite error term:
\(\nu_{it} = \alpha_{i} + \varepsilon_{it}\) and rewrite the model in
Equation \eqref{eq:gpm} as \begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
\end{align} where
\(\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}\) and
\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones. The strict
exogeneity assumption in the RE model implies \begin{align}
\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
\mathop{\mathrm{\mathbb{E}}}[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] & = 0,
\end{align} where \(\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}\). For
both parts, the assumption that the unconditional expectations are 0 is
unproblematic as long as a constant is included in the regression. The
first part says the idiosyncratic errors at each timepoint are assumed
to be independent of the explanatory variables at \textit{all}
timepoints which is stronger than just assuming that they are
\textit{contemporaneously} independent. This implies that they are also
uncorrelated, i.e.,
\(\mathop{\mathrm{\mathbb{E}}}[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\)
and
\(\mathop{\mathrm{\mathbb{E}}}[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
\citep{Wooldridge2002, Bruederl2015}. We assume the idiosyncratic errors
are further independent of the individual effects, which implies
\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}\varepsilon_{it}] = 0\).

The second part is the potentially controversial assumption: it states
that the individual effects are uncorrelated with the independent
variables. We can use an intuitive concrete example to show why this is
often controversial: If we are interested in the question of whether
married men earn more than unmarried men, then the second part of the
strict exogeneity assumption means that a man's marriage status would
have to be uncorrelated with all the time-invariant characteristics that
could potentially make that man an attractive marriage candidate in the
first place; e.g., looks, personality, family's status, profession, etc.
\citep{Bruederl2015}.\footnote{Assuming, for the sake of argument, that
  these characteristics are constant over time.}

From what we have discussed so far, the \(T \times T\) covariance matrix
of the errors
\(\bm{\Omega}_{i} = \mathop{\mathrm{\mathbb{E}}}[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\)
can be constructed. However, the standard random effects model adds the
additional assumptions \begin{alignat}{3}
\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
\end{alignat} i.e., the idiosyncratic errors are conditionally
homoscedastic and serially uncorrelated and \begin{alignat}{2}
\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
\end{alignat} i.e., the individual effects are conditionally
homoscedastic (they are necessarily serially correlated as long as
\(\sigma_{\alpha}^{2} > 0\)). From that, we arrive at the typical random
effects structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):

\begin{align}
\bm{\Omega} & = 
\begin{pmatrix}
\bm{\Omega}_{1} & \hdots & 0 & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & \bm{\Omega}_{i} & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & 0 & \hdots & \bm{\Omega}_{N}
\end{pmatrix}
\end{align} with \(T \times T\) typical elements \begin{align}
\bm{\Omega}_{i} & = 
\begin{pmatrix}
\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
\end{pmatrix} \label{eq:omeganui}
\end{align} where
\(\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}\)
\citep{Wooldridge2002, Schmidheiny2019}. This means that in the
conditional covariance matrix of the errors, given the time-varying and
-invariant covariates, units over time will be correlated due to the
individual effects. We should keep the covariance structure of the
errors in mind as it will help make sense of the use of latent variables
to decompose the dependent variable into between- and within-variance
components, discussed below in Section \ref{fe-sem}.

Estimation of the RE model can be done using feasible generalized least
squares (GLS) in which the two unknowns in \(\bm{\Omega}\),
\(\sigma_{\alpha}^{2}\) and \(\sigma_{\nu}^{2}\), are first estimated
using pooled ordinary least squares (pooled OLS or POLS),
where\footnote{Normally a degrees-of-freedom correction is applied by
  subtracting off the number of independent variables that is negligible
  in large N samples. It is ignored here for the sake of simplicity.}
\begin{align}
\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
\end{align} and
\(\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}\),
\(\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}\) and
\(\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}\).
Then, the coefficients are estimated using those estimates in the
variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
\begin{pmatrix}
\hat{\bm{\beta}}_{RE} \\
\hat{\bm{\gamma}}_{RE}
\end{pmatrix} & = 
(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
\end{align} where
\(\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}\) and \(\bm{X}\)
is \(NT \times K\) and \(\bm{Z}\) is \(NT \times M\) and \(\bm{y}\) is
\(NT \times 1\).

In practice, however, computational problems can arise with large
cross-sectional samples, where it can become difficult to invert the
\(\hat{\bm{\Omega}}\) matrix. One solution is to use `partial-demeaning'
to transform the data before performing simple POLS: \begin{align}
(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
\end{align} where
\(\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}\),
and \(\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}\),
\(\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}\),
\(\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}\) and
\(\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}\)
\citep{R-plm_a}.

The RE model can also be estimated in the maximum likelihood framework,
where in the associated literature on panel models are generally
referred to as either mixed models, hierarchical models or longitudinal
models. The typical RE model discussed here is the equivalent to a mixed
model with random intercepts and fixed slopes \citep{R-plm_a}. Under the
assumption of normality, along with homoscedasticity and serially
uncorrelated errors, the maximum likelihood estimator is the same as the
OLS estimator. For more on the topic of mixed models, see for example
\citet{R-lme4}, \citet{Bates2010}.

\hypertarget{fe}{%
\subsection{Fixed effects}\label{fe}}

For fixed effects, we assume that the individual effects are
\textit{not independent} of the model covariates, i.e.,
\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] \ne 0\).
Under this assumption, grouping the individual effects in with the
composite error will cause the coefficients of interest, here
specifically \(\bm{\beta}\) to be inconsistent
\citep{Wooldridge2002, Wooldridge2012}. We write the model therefore
again in terms of the general panel model in Equation \eqref{eq:gpm}
with separate individual effects and idiosyncratic error terms. In order
to drop assumptions involving the individual effects, a number of
methods are available (e.g., differencing, least squares dummy variable
regression), but the most common approach is to \textit{demean} the
equation \citep{Bruederl2015}. Demeaning is the same as the
transformation applied in Equation \eqref{eq:partialdemean} in the
special case where \(\theta = 1\). I.e., demeaning involves subtracting
the per-unit, over-time average from each of the model terms, i.e.,
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
\end{align} where the over-time averages are calculated the same as
above, and the variables with the dots above them represent the demeaned
versions. Because the average of something that does not change is that
thing itself, the individual effects, along with any time-invariant
predictors, get wiped out by the demeaning. This means that no
assumptions about the relatedness of the model covariates and the
unit-specific portion of the error are needed. Consistency of the
estimates is related solely to the strict exogeneity assumption imposed
on the idiosyncratic errors, i.e.,
\(\mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}] = 0\)
which also implies
\(\mathop{\mathrm{\mathbb{E}}}[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
\citep{Bruederl2015, Wooldridge2002}.

Having demeaned the data, the typical FE estimator is POLS on the
transformed data \begin{align}
\bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
\end{align} \citep{Bruederl2015}. The downside to this approach is that
no time-invariant predictors can be included in the model. However,
there are alternative approaches in the random effects and mixed model
frameworks that allow them to be included. These models are sometimes
referred to as `within-between' or `hybrid' models, often based on the
Chamberlain \citeyearpar{Chamberlain1980} and Mundlak
\citeyearpar{Mundlak1978} approaches, see for example \citet{Bell2018};
\citet{Allison2011}; \citet{Schunck2013}; \citet{Enders2007}. In the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
appendix}, it will be discussed how to also get around this restriction
using SEM.

\hypertarget{fe-sem}{%
\section{Fixed effects in structural equation modeling}\label{fe-sem}}

Moving from the conventional methods outlined above to SEM, we must
state the FE model in a different way. We turn to latent variables to
account for time-invariant unobserved heterogeneity. In fact, besides
accounting for measurement error and the representation of abstract
hypothetical concepts, unobserved heterogeneity has historically been
one of the main uses of latent variables in SEM \citep{Skrondal2004}.

\hypertarget{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}{%
\subsection{Modeling time-invariant unobserved heterogeneity as a latent
variable}\label{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}}

We first need to convert the data from stacked, long-format vectors of
length \(NT\) into \(T\) individual vectors of length \(N\). To see why
this is necessary, consider what effect this has on the vector of
responses \(y_{it}\). Let us, for a minute ignore any covariates and
focus just on the dependent variable (a so-called `intercept-only' or
`null' model) so that we have
\(y_{it} = \alpha_{i} + \varepsilon_{it}\). When we convert the data to
wide-format, we get \(T\) individual equations, \begin{align}
\bm{y}_{t} & = \bm{\alpha} + \bm{\varepsilon}_{t} \\
\begin{bmatrix}
y_{1t} \\
y_{2t} \\
\vdots \\
y_{Nt}\end{bmatrix} & = 
\begin{bmatrix}
\alpha_{1} \\
\alpha_{2} \\
\vdots \\
\alpha_{N}
\end{bmatrix} + 
\begin{bmatrix}
\varepsilon_{1t} \\
\varepsilon_{2t} \\
\vdots \\
\varepsilon_{Nt}
\end{bmatrix} \label{eq:wide}
\end{align} for each \(t = 1, 2, ..., T\). Because the idiosyncratic
errors are assumed to be uncorrelated across units and across time, the
covariance between any two of the new wide vectors
\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{si}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}), \ t \ne s\).
Otherwise, when \(t = s\), the covariance
\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{ti}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}) + \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{ti})\).
This is the structure we saw above in a typical element of
\(\bm{\Omega}\).

And in fact this is exactly how a latent variable is used to account for
time-invariant unobserved heterogeneity. The dependent variable at each
timepoint is regressed onto the latent variable, see Figure
\ref{fig:fesem}. Here, the regression weights or `factor loadings' are
fixed to one to represent our assumption that the effect of the
time-invariant unobserved heterogeneity is constant over time.\footnote{In
  fact, the initial FE-SEM setup shown in the main article mimics the
  POLS methods described above in that it assumes constant effects and
  error variances over time. These assumptions can be loosened and
  tested, as will be shown in the
  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{supplementary
  materials}. For now, for the sake of simplicity and comparability, we
  retain the assumptions associated with the `pooled' models for the
  most part.} It also means that the estimated variance of the latent
variable is equal to the
\textit{average covariance between the wide-format columns of the dependent variable over time}.
If \(y_{it} = \alpha_{i} + \varepsilon_{it}\) is the true data
generating process, then the relationship between two units over time is
just \(\mathop{\mathrm{\mathrm{Var}}}(\alpha)\), regardless of the time
distance. Referring back to the random effects structure of
\(\bm{\Omega}_{i}\) in Equality \eqref{eq:omeganui} for a generic unit
\(i\), we see the covariance on all of the off-diagonals is
\(\sigma^{2}_{\alpha}\). And, as we know, the average of something that
does not change is that thing itself. I.e., if \(T(T-1)/2 = h\) is the
number of elements on either the upper- or lower triangle of
\(\bm{\Omega}_{i}\), then we have
\(h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}\).

To elaborate on this concept some more, consider the following matrix
equation of the variances and the nonredundant covariances in a
three-wave intercept-only model that follows directly from Equation
\eqref{eq:wide} (assuming
\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s\)),
and which we can solve easily with least squares: \begin{align}
\bm{A}\bm{x} & = \bm{b} \\
\begin{pmatrix}
1 & 1 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 1 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 1
\end{pmatrix}
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\mathop{\mathrm{\mathrm{Var}}}(y_{1}) \\
\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) \\
\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) \\
\mathop{\mathrm{\mathrm{Var}}}(y_{2}) \\
\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
\mathop{\mathrm{\mathrm{Var}}}(y_{3})
\end{pmatrix}
\end{align} where \(\psi = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\),
\(\phi_{t} = \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{t})\). We can
solve this equation to show \begin{align}
\bm{A}\bm{x} & = \bm{b} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\begin{pmatrix}
\hat{\psi} \\
\hat{\phi}_{1} \\
\hat{\phi}_{2} \\
\hat{\phi}_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
\mathop{\mathrm{\mathrm{Var}}}(y_{1}) - \hat{\psi} \\
\mathop{\mathrm{\mathrm{Var}}}(y_{2}) - \hat{\psi} \\
\mathop{\mathrm{\mathrm{Var}}}(y_{3}) - \hat{\psi}
\end{pmatrix}.
\end{align} So if, in fact the covariance between any two wide-format
columns of \(y\) is
\(\mathop{\mathrm{\mathrm{Cov}}}(y_{t},y_{s}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha), \ \forall \ s \ne t\),
then
\(\hat{\psi} = \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) = \frac{3 \mathop{\mathrm{\mathrm{Var}}}(\alpha)}{3} = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\).
This shows that if our assumption about the underlying DGP is correct,
i.e., \(y_{it} = \alpha_{i} + \varepsilon_{it}\) and
\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s\),
then the estimated variance of \(\alpha\) is just what it should be: the
average covariance between units of \(y\) over time. Once we add in
observed covariates, the estimated covariance of \(\alpha\) then become
the \textit{conditional} covariance of \(y\) over time, given those
covariates.

\hypertarget{model-notation}{%
\subsection{Model notation}\label{model-notation}}

Having explained how a latent variable is used to estimate the
individual effects, we need to state the FE model using SEM-compatible
matrix notation. There are a number of different model notations (see,
for example \citet{Bollen1989} for an overview), but the one that will
serve us best is one that was proposed by \citet{Graff1979}:
\begin{align}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
\end{align} where
\(\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}\),
\(\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}\),
\(\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}\). \(\bm{y}\) is a vector of
observed dependent variables and \(\bm{x}\) is a vector of observed
independent variables. \(\bm{\eta}\) is a vector of the latent dependent
variables and \(\bm{\xi}\) is a vector of latent independent variables.
\(\bm{\varepsilon}\) and \(\bm{\delta}\) are vectors of the errors of
the observed dependent and independent variables, respectively, and
\(\bm{\zeta}\) is a vector of the errors, or disturbances, of the latent
variables. Notice the \(^{+}\) symbol is just meant to differentiate the
vectors with them from those without them. That means, \(\bm{\eta}^{+}\)
is a vector that holds the observed and latent variables, both dependent
(in SEM they are referred to as `endogenous') and independent (i.e.,
`exogeneous')\footnote{For our purposes, the terms endogenous and
  dependent, on the one hand, and exogenous and independent, on the
  other, can be used interchangeably.}, \(\bm{\zeta}^{+}\) holds the
errors for the observed variables and the disturbances of the latent
variables. \(\bm{y}^{+}\) holds just the observed variables, both
dependent and independent, and \(\bm{\Lambda_{y}}^{+}\) is a matrix of
ones and zeros that selects the observed variables from
\(\bm{\eta}^{+}\). Lastly, \(\bm{B}\) is a matrix that holds the
regression coefficients.

If we say that \(p\) and \(q\) stand for the number of observed
dependent and independent variables, respectively, and \(m\) and \(n\)
stand for the number of latent dependent and independent variables,
respectively, then \(\bm{\eta}^{+}\) and \(\bm{\zeta}^{+}\) are
\(p + q + m + n\), \(\bm{y}^{+}\) is \(p + q\), \(\bm{\Lambda_{y}}^{+}\)
is \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
\((p + q + m + n) \times (p + q + m + n)\) \citep{Bollen1989}.

>>>>

% see https://stackoverflow.com/a/47122900

@@
